Chasing \rainbows and \oceanglints:\\ Inner working angle constraints for the Habitable Worlds Observatory

### 2023-05-25 src/tex/ms.tex

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (\HWO), a coronagraphic space mission to detect rocky planets in habitable zones, establish their habitability, and search them for biosignatures. 
%
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase curves of starlight reflected by an exoplanet can reveal ocean glint, rainbows and other phenomena caused by scattering by clouds or atmospheric gas.
%
Direct imaging missions are optimised for planets near quadrature, and \HWO' coronagraph may obscure the phase angles where such optical features are strongest. 
%
The range of accessible phase angles for a given exoplanet will depend on the planet's orbital inclination and/or the coronagraph's inner working angle (IWA). 
%
For planets in near edge-on orbits, accessible phase angles are limited by coronagraphic obscuration. 
%
We use \HWO'~catalogue of 160 target stars to estimate the number of exo-Earths that could be searched for ocean glint, rainbows, and Rayleigh scattering. 
%
We find that any exo-Earths in planetary systems in this catalogue will have an accessible Rayleigh scattering peak in their polarimetric phase curve. 
%
The rainbow due to water clouds at phase angles of $\sim\qtyrange{20}{60}{\degree}$ would be accessible to \HWO in $\sim\num{42}$ systems, while the ocean glint signature at phase angles of $\sim\qtyrange{130}{170}{\degree}$ would be accessible in $\sim\num{14}$ systems, assuming an \IWA of \qty{62}{\mas} ($3\,\lambda/D$ for a $D=\qty{6}{\meter}$ telescope at $\lambda=\qty{600}{\nano\meter}$).
%
Improving the \IWA to $2\,\lambda/D$ increases the number of systems with accessible rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases slightly for eccentric orbits.  %and scales approximately inversely with the \IWA.  
\end{abstract}

@@

Liquid surface water is closely related to planetary habitability because it is essential for life as we know it.
%
To first order, its presence can be predicted by the flux a planet receives from its star.
%
The habitable zone (HZ) is defined as the range of semi-major axes around a given main sequence star, for which a rocky planet may have the right temperature to maintain liquid surface water \citep{kasting93}. 
%
However, there are many factors besides the incident stellar flux that influence a planet's ability to hold liquid surface water, such as the atmospheric pressure and geological activity. 
%
Detecting liquid surface water on an exoplanet would thus provide important context for the interpretation of possible biosignatures.
%
Variations of the flux and degree of polarization as an exoplanet orbits its host star---phase variations---could reveal scattering phenomena from condensed water \citep[e.g., rainbows;][]{bailey2007} or specular reflection, i.e. glint, from water oceans \citep[][]{2008Icar..195..927W}.
%
However, depending on a planetary system's architecture, such features can be obscured by the inner working angle (\IWA) of a telescope's coronagraph, shown schematically in \cref{fig:annotated-orbit}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
    \centering
    \includegraphics{figures/figure-1-bott-plot.pdf}
    \script{create-figure-1-bott-plot.py}
    \caption{
        Phase curves of the normalised total and the normalised linearly polarised flux of a cloudy planet with a wavy ocean surface and an Earth-like atmosphere in an edge-on orbit (adapted from \citealt{treesandstam2019}).  
        The wind speed over the ocean---which determines the wave height \citep{CoxMunk1954}---is \qty{7}{\meter\per\second}, and the clouds consist of spherical water droplets.
        The cloud coverage fraction is 0.5, and while the cloud pattern is patchy, the glint is always cloud-free. 
        The fluxes are normalised such that the total (unpolarised + polarised) flux equals the planet's geometric albedo at phase angle $\alpha = \qty{0}{\degree}$.
        In the bottom panel, we illustrate the scattering angle ranges of glories, rainbows, Rayleigh scattering, and ocean glint. 
        Note that the features in the actual curves partly overlap with each other.
        The glory illustration is taken from \citet{Zakovryashin_2020}; all other illustrations are from \href{https://unsplash.com}{unsplash.com} [\href{https://unsplash.com/photos/YWL4dqqiRdk}{1}, \href{https://unsplash.com/photos/BgFOMcg4HNs}{2}, \href{https://unsplash.com/photos/9YdePM6BrO4}{3}].
        This figure is best viewed in colour.
        \looseness=-1
    }
    \label{fig:bottplot}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey \citep[Astro2020;][]{decadal} recommended the first in a new \enquote{Great Observatories} program a telescope with the capability to detect signatures of habitability on about 25 HZ planets.
%
This requires an instrument with a coronagraph capable of high-contrast imaging and low-resolution spectroscopy at optical to near-infrared wavelengths. 
%
The precursor technology recommended by the survey lists \enquote{direct imaging to probe polarized ocean glint on terrestrial planets} as a priority capability \citep[Box E.1 in][]{decadal} for ground and space-based observatories.
%
Following the release of this report, NASA recently announced the start of the development of the Habitable Worlds Observatory (hereafter \enquote{\HWO{}}).
%
The performance and characteristics (on- or off-axis, diameter, type of segmentation, polarimetry or photometry, and coronagraph, \IWA) of \HWO will significantly impact the expected exo-Earth yield \citep{Stark2019_exoplanetyield} and are still to be determined.
%
They will likely be heavily informed by the LUVOIR \citep{LUVOIR2019} and HabEx \citep{HabEx_2020} preparatory studies.
%
For generality, in this work, we report angular separations in milliarcseconds; however, for ease of comparison, \cref{tab:IWA_OWA} lists the angular separations corresponding to multiples of $\lambda / D$ for a $\qty{6}{\meter}$ telescope, as suggested in Astro2020, at representative wavelengths $\lambda$ of $\qty{600}{\nano\meter}$ and $\qty{1}{\micro\meter}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table 1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
    \centering
    \caption{
        Inner and Outer Working Angles (\IWA and \OWA{}, respectively) in milliarcseconds (\unit{\mas}) as functions of $\lambda / D$, at wavelengths $\lambda$ of \qty{600}{\nano\meter} and \qty{1}{\micro\meter}, assuming a telescope diameter $D$ of 6\,m.
    }
    \label{tab:IWA_OWA}
    \begin{tabular}{ c c c c } 
    \toprule
    & $\lambda/D$ & \unit{\mas} (at \qty{600}{\nano\meter}) & \unit{\mas} (at \qty{1}{\micro\meter}) \\
    \midrule
    \midrule
    IWA & 1 & 20.63 & 34.38 \\
    IWA & 2 & 41.25 & 68.75 \\
    IWA & 3 & 61.88 & 103.13 \\
    IWA & 4 & 82.51 & 137.51 \\
    \midrule
    OWA & 32 &  660.05 & 1100.08 \\
    OWA & 64 & 1320.09 & 2200.16 \\
    \bottomrule
    \end{tabular}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Outline}

In this paper, we quantify the planetary phase angles that would be accessible for direct observations of hypothetical terrestrial planets orbiting in the habitable zones of \HWO target stars.
%
In \cref{subsec:2.1}, we describe the target list, in \cref{subsec:2.2}, we derive expressions for the range of phase angles given a few different limits, in \cref{subsec:2.3}, we describe Monte Carlo simulations to marginalize over the unknown orbital inclination and eccentricity of planets around these stars, and in \cref{subsec:2.4}, we outline how the photometric and polarimetric planet/star contrast changes along a planet's orbit.  
%
We present and discuss our results in \cref{sec:3}, and conclude in \cref{sec:4}.

@@

A preliminary target list of nearby stars (maximum distance \qty{25}{\parsec}) around which \HWO may search for habitable rocky planets, assuming a \qty{6}{\meter} telescope, has been compiled: the \emph{NASA Exoplanet Exploration Program's Mission Star List for the Habitable Worlds Observatory}.%
\footnote{Available online: \url{https://exoplanets.nasa.gov/internal_resources/2645_NASA_ExEP_Target_List_HWO_Documentation_2023.pdf}}
%
This list consists of $\num{164}$ stars, the majority of which are Sun-like dwarfs: 66~F~dwarfs, 55~G~dwarfs, 40~K~dwarfs, and 3~M~dwarfs.
%
The target list is constructed assuming a maximum planet magnitude in the Cousins Rc band of $R_c = 31$ and a lower limit of \num{2.5e-11} for the planet-star flux ratio. 
% 
The adopted HZ limits are a semi-major axis of \qtyrange{0.95}{1.67}{\au} for a solar twin and planet sizes between \qtyrange{0.8}{1.4}{} Earth radii. 
%
For non-solar stars, the semi-major axis scales as the square root of the bolometric luminosity normalized to the Sun. 
%
This range of orbital separation corresponds to the \enquote{conservative habitable zone} \citep{kasting93, kopparapu13}. 
%
In the remainder of this work, we consider hypothetical Earth-sized planets orbiting at the right distance from their parent star to receive the same flux as the Earth receives from the Sun.
%
By design, such planets should be observable by \HWO at quadrature phase.

@@

\Cref{fig:scatterplot} shows $\Delta \alpha$ for the systems in the \HWO target list, assuming an IWA of \qty{62}{\mas}.
%
As can be seen in \cref{fig:scatterplot}, for exoplanets in edge-on circular orbits, it will always be possible to observe the peak polarization due to Rayleigh scattering, which occurs around quadrature ($\alpha=\qty{90}{\degree}$), while the rainbow angle (around $\alpha=\qty{40}{\degree}$) will only be accessible for a subset of exoplanets.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 4
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
    \centering
    \includegraphics{figures/figure-4-scatterplot.pdf}
    \script{create-figure-4-scatterplot.py}
    \caption{
        Scatter plot for the \HWO target stars, showing stellar effective temperatures and distances.
        %
        The circle sizes represent the angular separation between the star and its planet (in mas) as presented in the target list.
        %
        The colours show $\Delta \alpha$ for hypothetical planets on circular, edge-on orbits at a semi-major axis $a$ corresponding to an Earth-like instellation and an IWA of \qty{62}{\mas}.
        %
        The colour bar indicates the optical phenomena that can in principle be detected: the phenomena from the bottom of the bar up to the colour of a given circle would be detectable for that planet.
        %
        For example, dark blue circles indicate planets for which most phenomena would be observable: planets that can be observed at the rainbow angle can also be observed at angles exhibiting Rayleigh scattering.
    }
    \label{fig:scatterplot}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 5
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}%[t]
    \centering
    \includegraphics{figures/figure-5-n-over-phase.pdf}
    \script{create-figure-5-n-over-phase.py}
    \caption{
        Cumulative distributions of the most extreme phase angles accessible for different \IWA and for randomly inclined, circular orbits (solid lines) and randomly inclined, eccentric orbits (dashed lines).
        %
        The top and bottom $x$-axes indicate the minimum and maximum accessible phase angles, respectively.
        %
        These angles are symmetric about quadrature (\qty{90}{\degree}).
        %
        The $y$-axis on the left indicates the number of planetary systems divided by the number of Monte Carlo samples, and is thus normalised to the number of systems on the target list.
        %
        The $y$-axis on the right indicates the number of HZ Earth-like planets that could be imaged at these phase angles, assuming an occurrence rate of $\eta_\oplus = \qty{24}{\percent}$.
        {\color{red}{the plot says 'elliptical orbits' instead of 'eccentric orbits'}}
    }
    \label{fig:betaallofit}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 7: Yarn ball
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
    \centering
    \includegraphics{figures/figure-7-yarnball.pdf}  
    \script{create-figure-7-yarnball.py}
    \caption{
        Examples of the eccentric orbits generated for the stellar sample. 
        Each plot shows 10 examples of orbits randomly drawn for each star.
        The orbits are scaled to ensure an Earth-like incident flux. 
        The concentric, dashed circles indicate inner working angles of \num{20.6}, \num{41.3}, \num{61.9}, and \qty{82.5}{\mas}.
        Each system has been scaled such that its HZ has the same radius, hence the variation in the values of the \IWA{}s compared to the orbits.
        It is clear that the \IWA can significantly affect the range of observable phases angles for each orbit.
    }
    \label{fig:ball-o-yarn}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We then calculate the minimum and maximum accessible phase angles for \IWA{}s of $\qtylist{21; 42; 63; 84}{\mas}$.
%
The dashed lines in \cref{fig:betaallofit} show the normalised cumulative distribution of the accessible phase angles for each value of the \IWA for eccentric orbits. 
%
We also calculate which target planets would be observable at the phase angles where ocean glint, the rainbow, or the Rayleigh peak occur.
\looseness=-1

@@

A second parameter that is crucial for detecting angular features like rainbows and ocean glint on Earth-like exoplanets is the contrast, that is, the planet/star flux ratio. 
%
While contrast is not the main focus of this paper, we note that the total and polarized flux contrasts depend on the phase angle, due to a combination of the changing illumination fraction of the observable planetary disk and non-isotropic reflection by the planet.
%
For the contrast calculations we start with the NASA Exoplanet Exploration Program’s Mission Star List for \HWO, which includes for every star an estimated contrast for an Earth-twin in the habitable zone.
%
The contrast is given by \citep[e.g.,][]{2023A&A...671A.165M}
\begin{equation}
    C(\alpha) 
    = F_p(\alpha) / F_* 
    = A_\mathrm{G} \cdot \phi(\alpha) \cdot \left( R_p / a(\alpha) \right)^2 \,,
    \label{eq:contrast}
\end{equation}
where $A_\mathrm{G}$ is the planet's geometric albedo, $R_p$ its radius, $\phi(\alpha)$ is the phase function at phase angle $\alpha$ (normalized to \num{1.0} at full phase, thus at $\alpha=\qty{0}{\degree}$), and $a$ is the distance between the planet and its host star.
%
The contrast estimations for \HWO assumed circular orbits and a Lambertian phase function  \citep{Russell1916} with $A_\mathrm{G}=0.2$. 
%
For a more accurate representation of the contrast as function of separation we can update the phase function with the models shown in Fig. \Cref{fig:bottplot} from \citet{treesandstam2019}. 
%
To update the phase function we will first correct for the reflected flux at quadrature and then map the normalise total flux and polarized flux as function of separation.

%
We assume that the calculated contrasts in the Star List for \HWO correspond to the reflected flux at quadrature and use \cref{eq:contrast} to calculate the reflected light fraction at $\alpha=\qty{90}{\degree}$ as $C(90) = \qty{6.37}{\percent}$. 
%
In addition, we calculate the reflected light fraction at quadrature with the updated model of \citet{treesandstam2019}, which assumes an Earth-like planet with an ocean surface, a wind-speed of $\qty{7}{\meter\per\second}$, and patchy liquid water clouds, observed at $\qty{670}{\nano\meter}$.
%
We correct the star list contrast by dividing by the \qty{6.37}{\percent} reflected light fraction of the Lambertian phase function and multiplying by the reflected light fraction of the updated model. 
%
We map the orbital phase for circular orbits to the phase angle and on-sky separation, given an inclination $i$ of \qty{90}{\degree}. 
%
Together, they give the reflected light contrast as a function of separation for a full orbit.
%
We multiply this flux with the degree of polarization to obtain the polarized contrast.
%
\Cref{fig:contrasts} shows the resulting contrast curves for three of the closest target stars: the M~dwarf Lalande~21185, the K~dwarf $\epsilon$~Eri, and the G~dwarf $\alpha$~Cen~A
%
These calculations are by no means complete or exact, given the many variables that go into the modelling of reflected total and polarized fluxes of Earth-like exoplanets \citep{treesandstam2019,trees2022}.

>>>>

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (\HWO), a coronagraphic space mission to detect rocky planets in habitable zones, establish their habitability, and search them for biosignatures. 
%
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase curves of starlight reflected by an exoplanet can reveal ocean glint, rainbows and other phenomena caused by scattering by clouds or atmospheric gas.
%
Direct imaging missions are optimised for planets near quadrature, but \HWO' coronagraph may obscure the phase angles where such optical features are strongest. 
%
The range of accessible phase angles for a given exoplanet will depend on the planet's orbital inclination and/or the coronagraph's inner working angle (IWA). 
%
For planets in near edge-on orbits, accessible phase angles are limited by coronagraphic obscuration. 
%
We use \HWO'~catalogue of 164 target stars to estimate the number of exo-Earths that could be searched for ocean glint, rainbows, and Rayleigh scattering. 
%
We find that most exo-Earths planetary systems in this catalogue will be observable at phase angles at which the Rayleigh scattering peak occurs in their polarimetric phase curve. 
%
The rainbow due to water clouds at phase angles of $\sim\qtyrange{20}{60}{\degree}$ would be observable with \HWO in $\sim\num{46}$ systems, while the ocean glint signature at phase angles of $\sim\qtyrange{130}{170}{\degree}$ would be observable in $\sim\num{16}$ systems, assuming an \IWA of \qty{62}{\mas} ($3\,\lambda/D$ for a $D=\qty{6}{\meter}$ telescope at $\lambda=\qty{600}{\nano\meter}$).
%
Improving the \IWA to $2\,\lambda/D$ increases the number of systems with observable rainbows and glints by factors of approximately 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases slightly for eccentric orbits.
%
By observing these scattering features, \HWO could detect a surface ocean and water cycle, key indicators of habitability.
\looseness=-1
\end{abstract}

@@

Liquid surface water is closely related to planetary habitability because it is essential for life as we know it.
%
To first order, its presence can be predicted by the flux a planet receives from its star.
%
The habitable zone (HZ) is defined as the range of semi-major axes around a given main sequence star, for which a rocky planet may have the right temperature to maintain liquid surface water \citep{kasting93}. 
%
However, there are many factors besides the incident stellar flux that influence a planet's ability to hold liquid surface water, such as the atmospheric pressure and geological activity. 
%
Detecting liquid surface water on an exoplanet would thus provide important context for the interpretation of possible biosignatures.
%
Variations of the flux and degree of polarization as an exoplanet orbits its host star---phase variations---could reveal scattering phenomena from condensed water \citep[e.g., rainbows;][]{bailey2007} or specular reflection (i.e., glint) from water oceans \citep[][]{2008Icar..195..927W}.
%
However, depending on a planetary system's architecture, such features can be obscured by the inner working angle (\IWA) of a telescope's coronagraph, shown schematically in \cref{fig:annotated-orbit}.
%
In this paper, we quantify the planetary phase angles that would be accessible for direct observations of hypothetical terrestrial planets orbiting in the habitable zones of \HWO target stars.
\looseness=-1

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
    \centering
    \includegraphics{figures/figure-1-bott-plot.pdf}
    \script{create-figure-1-bott-plot.py}
    \caption{
        Phase curves of the normalised total and normalised linearly polarised flux of a cloudy planet with a wavy ocean surface and an Earth-like atmosphere in an edge-on orbit (adapted from \citealt{treesandstam2019}).  
        The wind speed over the ocean---which determines the wave height \citep{CoxMunk1954}---is \qty{7}{\meter\per\second}, and the clouds consist of spherical water droplets.
        The cloud coverage fraction is 0.5, and while the cloud pattern is patchy, the glint is always cloud-free. 
        The fluxes are normalised such that the total (i.e., unpolarised + polarised) flux equals the planet's geometric albedo at phase angle $\alpha = \qty{0}{\degree}$.
        In the bottom panel, we illustrate the scattering angle ranges of glories, rainbows, Rayleigh scattering, and ocean glint. 
        Note that the features in the actual curves partly overlap.
        The glory illustration is taken from \citet{Zakovryashin_2020}; all other illustrations are from \href{https://unsplash.com}{unsplash.com} [\href{https://unsplash.com/photos/YWL4dqqiRdk}{1}, \href{https://unsplash.com/photos/BgFOMcg4HNs}{2}, \href{https://unsplash.com/photos/9YdePM6BrO4}{3}].
        This figure is best viewed in colour.
    }
    \label{fig:bottplot}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey \citep[Astro2020;][]{decadal} recommended a telescope with the capability to detect signatures of habitability on approximately 25 HZ planets in a new \enquote{Great Observatories} program.
%
This requires an instrument with a coronagraph capable of high-contrast imaging and low-resolution spectroscopy at optical to near-infrared wavelengths. 
%
The precursor technology recommended by the survey lists \enquote{direct imaging to probe polarized ocean glint on terrestrial planets} as a priority capability \citep[Box E.1 in][]{decadal} for ground and space-based observatories.
%
Following the release of this report, NASA recently announced the start of the development of the Habitable Worlds Observatory.
%
The performance and characteristics (e.g. on- or off-axis, diameter, type of segmentation, polarimetry or photometry, and coronagraph, \IWA) of \HWO will significantly impact the expected exo-Earth yield \citep{Stark2019_exoplanetyield} and are still to be determined.
%
They will likely be heavily informed by the LUVOIR \citep{LUVOIR2019} and HabEx \citep{HabEx_2020} preparatory studies.
%
For generality, in this work, we report angular separations in milliarcseconds; however, for ease of comparison, \cref{tab:IWA_OWA} lists the angular separations corresponding to multiples of $\lambda / D$ for a $\qty{6}{\meter}$ telescope, as suggested in Astro2020, at representative wavelengths $\lambda$ of $\qty{600}{\nano\meter}$ and $\qty{1}{\micro\meter}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table 1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
    \centering
    \caption{
        Inner and Outer Working Angles (\IWA and \OWA{}, respectively) in milliarcseconds (\unit{\mas}) as functions of $\lambda / D$, at wavelengths $\lambda$ of \qty{600}{\nano\meter}, \qty{670}{\nano\meter}, and \qty{1}{\micro\meter}, assuming a telescope diameter $D$ of \qty{6}{\meter}.
    }
    \label{tab:IWA_OWA}
    \script{create-table-1-iwa-values.py}
    \input{tables/table-1-iwa-owa}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \subsection{Outline}

% In this paper, we quantify the planetary phase angles that would be accessible for direct observations of hypothetical terrestrial planets orbiting in the habitable zones of \HWO target stars.
%
% In \cref{subsec:2.1}, we describe the target list, in \cref{subsec:2.2}, we derive expressions for the range of phase angles given a few different limits, in \cref{subsec:2.3}, we describe Monte Carlo simulations to marginalize over the unknown orbital inclination and eccentricity of planets around these stars, and in \cref{subsec:2.4}, we outline how the photometric and polarimetric planet/star contrast changes along a planet's orbit.  
%
% We present and discuss our results in \cref{sec:3}, and conclude in \cref{sec:4}.

@@

A preliminary target list of nearby stars (maximum distance \qty{25}{\parsec}) around which \HWO may search for habitable rocky planets, assuming a \qty{6}{\meter} telescope, has been compiled: the \emph{NASA Exoplanet Exploration Program's Mission Star List for the Habitable Worlds Observatory}.%
\footnote{Available online: \url{https://exoplanets.nasa.gov/internal_resources/2645_NASA_ExEP_Target_List_HWO_Documentation_2023.pdf}}
%
This list consists of $\num{164}$ stars, the majority of which are Sun-like dwarfs: 66~F~dwarfs, 55~G~dwarfs, 40~K~dwarfs, and 3~M~dwarfs.
%
The target list is constructed assuming a maximum planet magnitude in the Cousins $R_c$ band of $31$ mag and a lower limit of \num{2.5e-11} for the planet-star flux ratio. 
% 
The adopted HZ limits are a semi-major axis of \qtyrange{0.95}{1.67}{\au} for a solar twin and planet sizes between \qtyrange{0.8}{1.4}{} Earth radii. 
%
For non-solar stars, the semi-major axis scales as the square root of the bolometric luminosity normalized to the Sun. 
%
This range of orbital separation corresponds to the \enquote{conservative habitable zone} \citep{kasting93, kopparapu13}. 
%
In the remainder of this work, we consider hypothetical Earth-sized planets orbiting at the right distance from their parent star to receive the same flux as the Earth receives from the Sun.
%
By design, all such planets should be observable by \HWO at quadrature phase. 
\looseness=-1

@@

\Cref{fig:scatterplot} shows $\Delta \alpha$ for the systems in the \HWO target list, assuming an IWA of \qty{62}{\mas}.
%
As can be seen in \cref{fig:scatterplot}, for exoplanets in edge-on circular orbits, it will always be possible to observe the peak polarization due to Rayleigh scattering, which occurs around quadrature ($\alpha=\qty{90}{\degree}$), while the rainbow angle ($\alpha\sim\qty{40}{\degree}$) will only be accessible for a subset of exoplanets.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 4
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
    \centering
    \includegraphics{figures/figure-4-scatterplot.pdf}
    \script{create-figure-4-scatterplot.py}
    \caption{
        Scatter plot for the \HWO target stars, showing stellar effective temperatures and distances.
        %
        %The circle sizes represent the angular separation between the star and its planet (in mas) as presented in the target list.
        %
        The colours show $\Delta \alpha$ for hypothetical planets on circular, edge-on orbits at a semi-major axis $a$ corresponding to an Earth-like instellation and an IWA of \qty{62}{\mas}.
        %
        The colour bar indicates the optical phenomena that can in principle be detected: all the phenomena from the bottom of the bar up to the colour of a given circle would be detectable for that planet.
        %
        For example, dark blue circles indicate planets for which most phenomena would be observable: planets that can be observed at the rainbow angle can also be observed at angles exhibiting Rayleigh scattering.
    }
    \label{fig:scatterplot}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 5
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}%[t]
    \centering
    \includegraphics{figures/figure-5-n-over-phase.pdf}
    \script{create-figure-5-n-over-phase.py}
    \caption{
        Cumulative distributions of the most extreme phase angles accessible for different \IWA and for randomly inclined, circular orbits (solid lines) and randomly inclined, eccentric orbits (dashed lines).
        %
        The top and bottom $x$-axes indicate the minimum and maximum accessible phase angles, respectively.
        %
        These angles are symmetric about quadrature (\qty{90}{\degree}).
        %
        The $y$-axis on the left indicates the number of planetary systems divided by the number of Monte Carlo samples, and is thus normalised to the number of systems on the target list.
        %
        The $y$-axis on the right indicates the number of HZ Earth-like planets that could be imaged at these phase angles, assuming an occurrence rate of $\eta_\oplus = \qty{24}{\percent}$.
        %{\color{red}{the plot says 'elliptical orbits' instead of 'eccentric orbits'}}
    }
    \label{fig:betaallofit}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 6: Yarn ball
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
    \centering
    \includegraphics{figures/figure-6-yarnball.pdf}  
    \script{create-figure-6-yarnball.py}
    \caption{
        Examples of the eccentric orbits generated for the stellar sample. 
        Each plot shows 10 examples of orbits randomly drawn for each star.
        The orbits are scaled to ensure an Earth-like incident flux. 
        The concentric, dashed circles indicate inner working angles of \num{20.6}, \num{41.3}, \num{61.9}, and \qty{82.5}{\mas}.
        Each system has been scaled such that its HZ has the same radius, hence the variation in the values of the \IWA{}s compared to the orbits.
        It is clear that the \IWA can significantly affect the range of observable phases angles for each orbit.
    }
    \label{fig:ball-o-yarn}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We then calculate the minimum and maximum accessible phase angles for \IWA{}s of $\qtylist{21; 41; 62; 83}{\mas}$.
%
The dashed lines in \cref{fig:betaallofit} show the normalised cumulative distribution of the accessible phase angles for each value of the \IWA for eccentric orbits. 
%
We also calculate which target planets would be observable at the phase angles where ocean glint, the rainbow, or the Rayleigh peak occur.
\looseness=-1

@@

A second parameter that is crucial for detecting angular features like rainbows and ocean glint on Earth-like exoplanets is the contrast, that is, the planet/star flux ratio. 
%
While contrast is not the main focus of this paper, we note that the total and polarized flux contrasts depend on the phase angle, due to a combination of the changing illumination fraction of the observable planetary disk and non-isotropic reflection by the planet.
%
For the contrast calculations, we start with the NASA Exoplanet Exploration Program’s Mission Star List for \HWO, which includes for every star an estimated contrast for an Earth-twin in the habitable zone.
%
The contrast is given by \citep[e.g.,][]{2023A&A...671A.165M}
\begin{equation}
    C(\alpha) 
    = F_p(\alpha) / F_* 
    = A_\mathrm{G} \cdot \phi(\alpha) \cdot \left( R_p / a(\alpha) \right)^2 \,,
    \label{eq:contrast}
\end{equation}
where $A_\mathrm{G}$ is the planet's geometric albedo, $R_p$ its radius, $\phi(\alpha)$ is the phase function at phase angle $\alpha$ (normalized to \num{1.0} at full phase, thus at $\alpha=\qty{0}{\degree}$), and $a$ is the distance between the planet and its host star.
%
The contrast estimations for \HWO assumed circular orbits and a Lambertian phase function  \citep{Russell1916} with $A_\mathrm{G}=0.2$. 
%
For a more accurate representation of the contrast as a function of separation, we can update the phase function with the models shown in Fig. \Cref{fig:bottplot} from \citet{treesandstam2019}. 
%
To update the phase function, we will first correct for the reflected flux at quadrature and then map the normalised total flux and polarized flux as a function of separation.

%
We assume that the calculated contrasts in the Star List for \HWO correspond to the reflected flux at quadrature and use \cref{eq:contrast} to calculate the reflected light fraction at $\alpha=\qty{90}{\degree}$ as $C(90) = \qty{6.37}{\percent}$. 
%
In addition, we calculate the reflected light fraction at quadrature with the updated model of \citet{treesandstam2019}, which assumes an Earth-like planet with an ocean surface, a wind-speed of $\qty{7}{\meter\per\second}$, and patchy liquid water clouds, observed at $\qty{670}{\nano\meter}$.
%
We correct the star list contrast by dividing by the \qty{6.37}{\percent} reflected light fraction of the Lambertian phase function and multiplying by the reflected light fraction of the updated model. 
%
We map the orbital phase for circular orbits to the phase angle and on-sky separation, given an inclination $i$ of \qty{90}{\degree}. 
%
Together, they give the reflected light contrast as a function of separation for a full orbit.
%
We multiply this flux with the degree of polarization to obtain the polarized contrast.
%
\Cref{fig:contrasts} shows the resulting contrast curves for three of the closest target stars: the M~dwarf Lalande~21185, the K~dwarf $\epsilon$~Eri, and the G~dwarf $\alpha$~Cen~A.
%
These calculations are by no means complete or exact, given the many variables that go into the modelling of reflected total and polarized fluxes of Earth-like exoplanets \citep{treesandstam2019,trees2022}.
